---
title: "Used Car Price Prediction: Analysis & Modeling"
author: "Aditya Arte"
date: "2025-02-13"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

# 1. Introduction

This project analyzes a dataset of 10,000 used car listings to build a predictive model
for car prices. The goal is to understand which factors most influence used car pricing
and to develop an accurate, deployable prediction tool.

**Dataset:** [Kaggle - Car Price Dataset](https://www.kaggle.com/datasets/asinow/car-price-dataset/data)

# 2. Setup & Data Loading

```{r load-libraries}
library(tidyverse)
library(corrplot)
library(caret)
library(scales)
library(gridExtra)
```

```{r load-data}
car_data <- read.csv("car_price_dataset.csv")
```

# 3. Exploratory Data Analysis

## 3.1 Data Overview

```{r data-structure}
str(car_data)
```

```{r data-summary}
summary(car_data)
```

```{r data-head}
head(car_data, 10)
```

## 3.2 Data Quality

```{r null-check}
cat("Missing values per column:\n")
colSums(is.na(car_data))
```

```{r duplicate-check}
cat("Number of duplicate rows:", sum(duplicated(car_data)), "\n")
```

```{r unique-counts}
cat("Unique values per categorical column:\n")
cat("  Brands:", n_distinct(car_data$Brand), "\n")
cat("  Models:", n_distinct(car_data$Model), "\n")
cat("  Fuel Types:", n_distinct(car_data$Fuel_Type), "\n")
cat("  Transmissions:", n_distinct(car_data$Transmission), "\n")
cat("  Door configs:", n_distinct(car_data$Doors), "\n")
cat("  Owner counts:", n_distinct(car_data$Owner_Count), "\n")
```

The dataset is clean with no missing values or duplicates. This is consistent with a
well-curated dataset.

## 3.3 Target Variable: Price Distribution

```{r price-distribution, fig.height=5}
p1 <- ggplot(car_data, aes(x = Price)) +
  geom_histogram(binwidth = 1000, fill = "#2c3e50", alpha = 0.8, color = "white") +
  geom_vline(aes(xintercept = mean(Price)), color = "#e74c3c", linetype = "dashed", size = 1) +
  annotate("text", x = mean(car_data$Price) + 2000, y = Inf, vjust = 2,
           label = paste("Mean:", dollar(mean(car_data$Price))),
           color = "#e74c3c", fontface = "bold") +
  labs(title = "Distribution of Car Prices",
       subtitle = "Red dashed line indicates the mean price",
       x = "Price ($)", y = "Count") +
  scale_x_continuous(labels = dollar) +
  theme_minimal(base_size = 13)

p2 <- ggplot(car_data, aes(x = Price)) +
  geom_density(fill = "#3498db", alpha = 0.5) +
  labs(title = "Price Density", x = "Price ($)", y = "Density") +
  scale_x_continuous(labels = dollar) +
  theme_minimal(base_size = 13)

grid.arrange(p1, p2, ncol = 2)
```

Prices are approximately normally distributed with a slight right skew, indicating a
small number of higher-priced vehicles.

## 3.4 Numerical Features

```{r numerical-distributions, fig.height=8}
p_year <- ggplot(car_data, aes(x = Year)) +
  geom_histogram(binwidth = 1, fill = "#1abc9c", alpha = 0.8, color = "white") +
  labs(title = "Manufacturing Year", x = "Year", y = "Count") +
  theme_minimal(base_size = 12)

p_engine <- ggplot(car_data, aes(x = Engine_Size)) +
  geom_histogram(binwidth = 0.2, fill = "#9b59b6", alpha = 0.8, color = "white") +
  labs(title = "Engine Size", x = "Engine Size (L)", y = "Count") +
  theme_minimal(base_size = 12)

p_mileage <- ggplot(car_data, aes(x = Mileage)) +
  geom_histogram(binwidth = 10000, fill = "#e67e22", alpha = 0.8, color = "white") +
  labs(title = "Mileage", x = "Mileage (km)", y = "Count") +
  scale_x_continuous(labels = comma) +
  theme_minimal(base_size = 12)

p_owners <- ggplot(car_data, aes(x = factor(Owner_Count))) +
  geom_bar(fill = "#e74c3c", alpha = 0.8) +
  labs(title = "Owner Count", x = "Number of Previous Owners", y = "Count") +
  theme_minimal(base_size = 12)

grid.arrange(p_year, p_engine, p_mileage, p_owners, ncol = 2)
```

## 3.5 Categorical Features vs Price

```{r fuel-type-analysis, fig.height=5}
fuel_summary <- car_data %>%
  group_by(Fuel_Type) %>%
  summarise(
    Avg_Price = mean(Price),
    Median_Price = median(Price),
    Count = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Avg_Price))

p1 <- ggplot(car_data, aes(x = reorder(Fuel_Type, Price, FUN = median), y = Price, fill = Fuel_Type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  labs(title = "Price Distribution by Fuel Type", x = "Fuel Type", y = "Price ($)") +
  scale_y_continuous(labels = dollar) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p2 <- ggplot(fuel_summary, aes(x = reorder(Fuel_Type, -Avg_Price), y = Avg_Price, fill = Fuel_Type)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = dollar(round(Avg_Price))), vjust = -0.5, size = 3.5) +
  labs(title = "Average Price by Fuel Type", x = "Fuel Type", y = "Average Price ($)") +
  scale_y_continuous(labels = dollar) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

Electric and Hybrid vehicles command higher average prices than Diesel and Petrol cars.

```{r transmission-analysis, fig.height=5}
trans_summary <- car_data %>%
  group_by(Transmission) %>%
  summarise(
    Avg_Price = mean(Price),
    Median_Price = median(Price),
    Count = n(),
    .groups = "drop"
  )

p1 <- ggplot(car_data, aes(x = reorder(Transmission, Price, FUN = median), y = Price, fill = Transmission)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  labs(title = "Price by Transmission Type", x = "Transmission", y = "Price ($)") +
  scale_y_continuous(labels = dollar) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

p2 <- ggplot(trans_summary, aes(x = reorder(Transmission, -Avg_Price), y = Avg_Price, fill = Transmission)) +
  geom_col(alpha = 0.8) +
  geom_text(aes(label = dollar(round(Avg_Price))), vjust = -0.5, size = 3.5) +
  labs(title = "Average Price by Transmission", x = "Transmission", y = "Average Price ($)") +
  scale_y_continuous(labels = dollar) +
  scale_fill_brewer(palette = "Pastel1") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")

grid.arrange(p1, p2, ncol = 2)
```

Automatic vehicles tend to be priced higher than Manual and Semi-Automatic.

```{r brand-analysis, fig.height=6}
brand_summary <- car_data %>%
  group_by(Brand) %>%
  summarise(
    Avg_Price = mean(Price),
    Median_Price = median(Price),
    Count = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Avg_Price))

ggplot(brand_summary, aes(x = reorder(Brand, Avg_Price), y = Avg_Price, fill = Avg_Price)) +
  geom_col(alpha = 0.85) +
  geom_text(aes(label = dollar(round(Avg_Price))), hjust = -0.1, size = 3.2) +
  coord_flip() +
  labs(title = "Average Price by Brand",
       subtitle = "Higher-end brands command premium prices",
       x = "Brand", y = "Average Price ($)") +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_gradient(low = "#3498db", high = "#e74c3c") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

## 3.6 Key Relationships with Price

```{r scatter-plots, fig.height=10}
p1 <- ggplot(car_data, aes(x = Year, y = Price)) +
  geom_point(alpha = 0.15, color = "#2c3e50") +
  geom_smooth(method = "lm", color = "#e74c3c", se = TRUE) +
  labs(title = "Price vs Manufacturing Year", x = "Year", y = "Price ($)") +
  scale_y_continuous(labels = dollar) +
  theme_minimal(base_size = 12)

p2 <- ggplot(car_data, aes(x = Engine_Size, y = Price)) +
  geom_point(alpha = 0.15, color = "#2c3e50") +
  geom_smooth(method = "lm", color = "#e74c3c", se = TRUE) +
  labs(title = "Price vs Engine Size", x = "Engine Size (L)", y = "Price ($)") +
  scale_y_continuous(labels = dollar) +
  theme_minimal(base_size = 12)

p3 <- ggplot(car_data, aes(x = Mileage, y = Price)) +
  geom_point(alpha = 0.15, color = "#2c3e50") +
  geom_smooth(method = "lm", color = "#e74c3c", se = TRUE) +
  labs(title = "Price vs Mileage", x = "Mileage (km)", y = "Price ($)") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = dollar) +
  theme_minimal(base_size = 12)

p4 <- ggplot(car_data, aes(x = Owner_Count, y = Price)) +
  geom_point(alpha = 0.15, color = "#2c3e50", position = position_jitter(width = 0.2)) +
  geom_smooth(method = "lm", color = "#e74c3c", se = TRUE) +
  labs(title = "Price vs Owner Count", x = "Number of Owners", y = "Price ($)") +
  scale_y_continuous(labels = dollar) +
  theme_minimal(base_size = 12)

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

Key observations:

- **Year** has a strong positive correlation — newer cars are more expensive
- **Engine Size** shows a positive correlation — larger engines cost more
- **Mileage** has a negative correlation — higher mileage reduces value
- **Owner Count** shows minimal impact on price

## 3.7 Multi-Variable Visualization

```{r multi-variable, fig.height=7}
ggplot(car_data, aes(x = Mileage, y = Price, color = Fuel_Type, shape = Transmission)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_shape_manual(values = c("Manual" = 16, "Automatic" = 4, "Semi-Automatic" = 17)) +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = dollar) +
  labs(title = "Price vs Mileage by Fuel Type & Transmission",
       subtitle = "Electric/Hybrid vehicles (green/blue) cluster at higher prices",
       x = "Mileage (km)", y = "Price ($)",
       color = "Fuel Type", shape = "Transmission") +
  theme_minimal(base_size = 13)
```

```{r year-color, fig.height=6}
ggplot(car_data, aes(x = Mileage, y = Price, color = Year)) +
  geom_point(alpha = 0.5, size = 1.5) +
  scale_color_viridis_c(option = "plasma") +
  scale_x_continuous(labels = comma) +
  scale_y_continuous(labels = dollar) +
  labs(title = "Price vs Mileage Colored by Manufacturing Year",
       subtitle = "Newer vehicles (lighter colors) cluster at higher prices",
       x = "Mileage (km)", y = "Price ($)", color = "Year") +
  theme_minimal(base_size = 13)
```

## 3.8 Correlation Analysis

```{r correlation-matrix, fig.height=7}
# Encode categorical variables numerically for correlation
encoded_data <- car_data %>%
  mutate(
    Fuel_Type_Num = as.numeric(as.factor(Fuel_Type)),
    Transmission_Num = as.numeric(as.factor(Transmission)),
    Brand_Num = as.numeric(as.factor(Brand)),
    Model_Num = as.numeric(as.factor(Model))
  )

numeric_cols <- encoded_data %>%
  select(Price, Year, Engine_Size, Mileage, Doors, Owner_Count,
         Fuel_Type_Num, Transmission_Num, Brand_Num, Model_Num)

cor_matrix <- cor(numeric_cols, use = "complete.obs")

corrplot(cor_matrix, method = "color", type = "upper",
         addCoef.col = "black", number.cex = 0.7,
         tl.col = "black", tl.srt = 45,
         col = colorRampPalette(c("#3498db", "white", "#e74c3c"))(200),
         title = "Feature Correlation Matrix",
         mar = c(0, 0, 2, 0))
```

**Key correlations with Price:**

- **Year** (+): Newer cars cost more
- **Engine Size** (+): Larger engines cost more
- **Mileage** (−): Higher mileage reduces price
- **Fuel Type** and **Transmission** show weaker but meaningful correlations

# 4. Feature Engineering

```{r feature-engineering}
# Calculate car age instead of using raw year
car_data$Car_Age <- max(car_data$Year) - car_data$Year

# Price per unit of engine size (for analysis only)
car_data$Price_Per_Liter <- car_data$Price / car_data$Engine_Size

cat("New features created:\n")
cat("  Car_Age: ranges from", min(car_data$Car_Age), "to", max(car_data$Car_Age), "years\n")
```

# 5. Model Building

## 5.1 Data Preparation

```{r data-prep}
# Select features for modeling — include ALL available features
model_data <- car_data %>%
  select(Brand, Model, Year, Engine_Size, Fuel_Type, Transmission,
         Mileage, Doors, Owner_Count, Price)

# Encode categorical variables as factors
model_data$Brand <- as.factor(model_data$Brand)
model_data$Model <- as.factor(model_data$Model)
model_data$Fuel_Type <- as.factor(model_data$Fuel_Type)
model_data$Transmission <- as.factor(model_data$Transmission)

# Train-test split (80/20)
set.seed(123)
train_index <- createDataPartition(model_data$Price, p = 0.8, list = FALSE)
train_data <- model_data[train_index, ]
test_data <- model_data[-train_index, ]

cat("Training set:", nrow(train_data), "rows\n")
cat("Test set:", nrow(test_data), "rows\n")
```

## 5.2 Linear Regression

```{r linear-regression}
lm_model <- train(
  Price ~ Brand + Model + Year + Engine_Size + Fuel_Type + Transmission + Mileage + Doors + Owner_Count,
  data = train_data,
  method = "lm",
  trControl = trainControl(method = "cv", number = 10)
)

lm_predictions <- predict(lm_model, newdata = test_data)

lm_mae <- mean(abs(lm_predictions - test_data$Price))
lm_rmse <- sqrt(mean((lm_predictions - test_data$Price)^2))
lm_r2 <- cor(test_data$Price, lm_predictions)^2

cat("Linear Regression Results:\n")
cat("  MAE:  $", round(lm_mae, 2), "\n")
cat("  RMSE: $", round(lm_rmse, 2), "\n")
cat("  R²:   ", round(lm_r2, 4), "\n")
```

## 5.3 Ridge Regression (Regularized)

```{r ridge-regression}
ridge_model <- train(
  Price ~ Brand + Model + Year + Engine_Size + Fuel_Type + Transmission + Mileage + Doors + Owner_Count,
  data = train_data,
  method = "ridge",
  trControl = trainControl(method = "cv", number = 10),
  preProcess = c("center", "scale")
)

ridge_predictions <- predict(ridge_model, newdata = test_data)

ridge_mae <- mean(abs(ridge_predictions - test_data$Price))
ridge_rmse <- sqrt(mean((ridge_predictions - test_data$Price)^2))
ridge_r2 <- cor(test_data$Price, ridge_predictions)^2

cat("Ridge Regression Results:\n")
cat("  MAE:  $", round(ridge_mae, 2), "\n")
cat("  RMSE: $", round(ridge_rmse, 2), "\n")
cat("  R²:   ", round(ridge_r2, 4), "\n")
```

## 5.4 Model Comparison

```{r model-comparison}
results <- data.frame(
  Model = c("Linear Regression", "Ridge Regression"),
  MAE = c(lm_mae, ridge_mae),
  RMSE = c(lm_rmse, ridge_rmse),
  R_Squared = c(lm_r2, ridge_r2)
)

results %>%
  arrange(RMSE) %>%
  knitr::kable(digits = 2, col.names = c("Model", "MAE ($)", "RMSE ($)", "R²"),
               caption = "Model Performance Comparison")
```

## 5.5 Model Diagnostics

```{r residual-analysis, fig.height=8}
# Use the linear regression model for diagnostics
residuals <- test_data$Price - lm_predictions

p1 <- ggplot(data.frame(Predicted = lm_predictions, Actual = test_data$Price),
             aes(x = Predicted, y = Actual)) +
  geom_point(alpha = 0.3, color = "#2c3e50") +
  geom_abline(slope = 1, intercept = 0, color = "#e74c3c", linetype = "dashed", size = 1) +
  labs(title = "Predicted vs Actual Prices",
       subtitle = "Points near the red line indicate accurate predictions",
       x = "Predicted Price ($)", y = "Actual Price ($)") +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  theme_minimal(base_size = 12)

p2 <- ggplot(data.frame(Predicted = lm_predictions, Residual = residuals),
             aes(x = Predicted, y = Residual)) +
  geom_point(alpha = 0.3, color = "#2c3e50") +
  geom_hline(yintercept = 0, color = "#e74c3c", linetype = "dashed") +
  labs(title = "Residual Plot",
       subtitle = "Residuals should be randomly scattered around zero",
       x = "Predicted Price ($)", y = "Residual ($)") +
  scale_x_continuous(labels = dollar) +
  theme_minimal(base_size = 12)

p3 <- ggplot(data.frame(Residual = residuals), aes(x = Residual)) +
  geom_histogram(binwidth = 10, fill = "#3498db", alpha = 0.7, color = "white") +
  geom_vline(xintercept = 0, color = "#e74c3c", linetype = "dashed") +
  labs(title = "Distribution of Residuals",
       subtitle = "Should be approximately normally distributed around zero",
       x = "Residual ($)", y = "Count") +
  theme_minimal(base_size = 12)

p4 <- ggplot(data.frame(Residual = residuals), aes(sample = Residual)) +
  stat_qq(alpha = 0.3, color = "#2c3e50") +
  stat_qq_line(color = "#e74c3c") +
  labs(title = "Q-Q Plot of Residuals",
       subtitle = "Points following the line suggest normality",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal(base_size = 12)

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

## 5.6 Feature Importance

```{r feature-importance, fig.height=5}
# Extract coefficients from the linear model (excluding individual model/brand levels)
lm_coefs <- summary(lm_model$finalModel)$coefficients
coef_df <- data.frame(
  Feature = rownames(lm_coefs),
  Estimate = lm_coefs[, "Estimate"],
  Std_Error = lm_coefs[, "Std. Error"],
  P_Value = lm_coefs[, "Pr(>|t|)"]
)

# Show the main numeric feature coefficients
main_features <- coef_df %>%
  filter(Feature %in% c("Year", "Engine_Size", "Mileage", "Doors", "Owner_Count")) %>%
  mutate(Significant = ifelse(P_Value < 0.05, "Yes", "No"))

ggplot(main_features, aes(x = reorder(Feature, abs(Estimate)), y = Estimate, fill = Significant)) +
  geom_col(alpha = 0.8) +
  coord_flip() +
  labs(title = "Linear Model Coefficients (Numeric Features)",
       subtitle = "Impact of each feature on predicted price",
       x = "Feature", y = "Coefficient Estimate ($)") +
  scale_fill_manual(values = c("Yes" = "#27ae60", "No" = "#95a5a6")) +
  theme_minimal(base_size = 13)
```

# 6. Save Final Model

```{r save-model}
# Save the best-performing model
saveRDS(lm_model, "car_price_model.rds")
cat("Model saved to car_price_model.rds\n")
```

# 7. Key Findings & Conclusions

1. **Price Drivers**: Manufacturing year, engine size, fuel type, and mileage are the
   strongest predictors of used car prices.

2. **Fuel Type Premium**: Electric and Hybrid vehicles command a price premium over
   traditional Diesel and Petrol vehicles, reflecting market trends toward green vehicles.

3. **Depreciation**: Each additional year of age reduces car value, while every additional
   kilometer of mileage decreases value — consistent with real-world depreciation patterns.

4. **Transmission Effect**: Automatic transmission vehicles are priced higher on average
   than Manual and Semi-Automatic options.

5. **Model Performance**: The linear regression model achieves strong performance metrics,
   indicating the selected features capture the key pricing dynamics in this dataset.

6. **Note on R² values**: The very high R² values observed suggest this dataset may have
   a synthetic or highly structured component. In production use, real-world data would
   likely show more noise and lower R² values (typically 0.75-0.90 for car pricing models).

---

*Analysis performed using R with tidyverse, caret, and corrplot packages.*
